---
title: "Mixed Poisson Example with `glmmBoot`"
author: "Daniel Flores Agreda"
date: 09-05-2020
categories: ["R"]
tags: ["glmmBoot", "Bootstrap"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE)
```

This code to illustrates the implementation of the bootstrapping methods described in [(Flores-Agreda and Cantoni, Under Review)](https://www.researchgate.net/publication/315768128_Bootstrapping_Generalized_Linear_Mixed_Models_via_a_Weighted_Laplace_Approximation) in a real-data context. Data comes from the \emph{Epilepsy} example found in [Molenberghs & Verbeke (2006)](https://www.springer.com/gp/book/9780387251448) initially reported by [Faught et. al. (1996)](https://www.ncbi.nlm.nih.gov/pubmed/8649570). 

## Analysis

To start performing the analysis, you need:

1. the following packages:
```{r, include = FALSE, message=FALSE, warnings=FALSE}
library(lme4)
library(glmmTMB)
library(sas7bdat)
library(TMB)
library(dplyr)
library(ggplot2)
library(kableExtra)
```

```{r, message=FALSE, warning = FALSE, echo = FALSE, include = FALSE}
compile(file = "~/Dropbox/Projects/Pro/glmmBoot/glmmBoot/src/glmm_rirs.cpp")
dyn.load(dynlib("~/Dropbox/Projects/Pro/glmmBoot/glmmBoot/src/glmm_rirs"))
source("~/Dropbox/Projects/Pro/glmmBoot/glmmBoot/R/glmmBoot.R")
epilepsy <- read.sas7bdat(file = "~/Dropbox/Projects/Pro/glmmBoot/glmmBoot/data/epilepsy.sas7bdat")
```

2. to compile `glmm_rirs.cpp` and load the resulting binary into the environment:
```r
compile(file = "glmmBoot/src/glmm_rirs.cpp")
dyn.load(dynlib("glmmBoot/src/glmm_rirs"))
```

3. the functions in the file `glmmBoot.R`:

```r
source("glmmBoot/R/glmmBoot.R")
```

4. the dataset `epilepsy.sas7bdat`
```r
epilepsy <- read.sas7bdat(file = "glmmBoot/data/epilepsy.sas7bdat")
```

## Model

The aim of the study was to verify the effects of an new anti-epileptic drug (AED) compared to a placebo on the number of seizures experienced by patients during the study. To do this, consider a mixed Poisson model for the outcome containing two potentially correlated random effects: one for a random intercept and another one for the visit time i.e. $y_{ij}|\mathbf{u}_i\sim \mathcal{P}(\lambda_{ij})$ with: 

$$\log(\lambda_{ij}) = \beta_0  + \beta_{1} T_{ij} + \beta_{2}t_{ij} + \beta_{3} T_{ij}t_{ij} +  \sigma_{1} u_{i1}  + \sigma_{2} u_{i2} t_{ij}$$

where:

- $T_{ij}$ represents the effect of the treatment and 
- $t_{ij}$ the visit time. 

The variance-covariance structure of the vector of Normal random effects $\mathbf{u}_i = [u_{i1}, u_{i2}]^T$, comprises a  correlation coefficient $\rho$ , i.e. 

$$\mathbf{D}(\mathbf{\sigma}) = \left[\begin{array}{c c} \sigma_{1}^2 & \rho\sigma_{1}\sigma_{2} \\  \rho\sigma_{1}\sigma_{2} & \sigma_{2}^2 \end{array}\right]$$.

## Implementation

### `estimate_glmm()`

The function `estimate_glmm()` estimates the model with a `TMB` template using the frame and estimates of objects of class `glmerMod` as starting values. For example: 
```{r}
obj.glmerMod <- lme4::glmer(nseizw~ trt*studyweek + (studyweek|id), 
                            family = "poisson",
                            data = epilepsy, 
                            control = glmerControl(optimizer = "bobyqa"))
```
for comparison purposes, let us also extract the estimates of this object
```{r, include = FALSE}
est_lmer <- extract_estimates(obj.glmerMod)
est_tmb  <- estimate_glmm(obj.glmerMod)
```
A comparison of the estimates yields:
```{r}
est_lmer %>% 
  inner_join(est_tmb, by=("Parameters"), suffix=c(" (glmer)", " (TMB)")) 
```

The generation of replicates using Random Weighted Laplace Bootstrap [(Flores-Agreda and Cantoni, Under Review)](https://www.researchgate.net/publication/315768128_Bootstrapping_Generalized_Linear_Mixed_Models_via_a_Weighted_Laplace_Approximation) is carried by the function `bootstrap_glmm()`, with the option `method=rwlb`. 
```{r}
## Bootstrap inference 
rwlb_reps <- bootstrap_glmm(obj.glmerMod,
                            B = 1000,
                            method = "rwlb")
```

The resulting Bootstrap replicates are stored in an object of the class `glmmBoot`, which can be used to construct 

- Estimates of the Parameters: by averaging over the replicates
- Estimates of the Standard Errors: by computing the standard deviations for the replicates and 
- Percentile-based Confidence Intervals (CI) with a level of 95%.

## Methods

### `confint()` 

produces bootstrap estimates, standard errors and confidence intervals with the `percentile` and `studentized` methods

- Example of the percentile method

```{r}
rwlb_reps %>% 
  confint(bootstrap_type = "percentile")
```

- Example of the studentized method
```{r}
rwlb_reps %>% 
  confint(bootstrap_type = "studentized")
```

### `plot()` 

Shows the boxplots of the replicates for all the parameters or a subset with a line describing the TMB estimates
    
- fixed effect parameters
```{r}
plot(rwlb_reps, 
     parm_subset=c("(Intercept)", "trt", "studyweek", "trt:studyweek")) + 
  ggtitle("Replicates of Fixed Effects") +
  theme_classic()
```

- variances of random effects 
```{r}
plot(rwlb_reps, 
     parm_subset=c("s_1", "s_2", "rho")) + 
  ggtitle("Replicates of Variance Components")+theme_classic()
```


